Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Swapped contexts for ite, tap

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      {\hasty{\Xi}{1}{a}{A}{\tint}}
      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} 
    \\
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\hasty{\Delta}{1}{e}{\mb{2}}{\tint}}
      {\haslb{\Xi}{s}{\ms{L}}}
      {\haslb{\Xi}{t}{\ms{L}}}
      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} 
    \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{0}{a}{A}{q}}
      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} 
    \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{0}{a}{A \otimes B}{q}}
      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \qquad
    \prftree[r]{\rle{where}}
      {\haslb{\Gamma}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L}{\ms{K}}}
      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
    \\
    \prftree[r]{\rle{nil-br}}
      {\lhaslb{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks}
  \Description{Typing rules for isotope-SSA blocks}
  \label{fig:ssa-block-typing}
\end{figure}

@@

\begin{figure}
  \begin{gather*}
    \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
      : \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})} \\
    \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}
      ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
    \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
      ; \dnt{\hasty{\Delta}{1}{e}{\mb{2}}{\tint}} \otimes \dnt{\Xi})}
      ; \ms{ite}_{\dnt{\Xi}} ;
      \\ \qquad \qquad \qquad \qquad
      [
        \dnt{\haslb{\Xi}{s}{\ms{L}}}, 
        \dnt{\haslb{\Xi}{t}{\ms{L}}} 
      ]
      \\
    \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A}{q}}
      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
      \\
    \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A \otimes B}{q}}
      ; \alpha
      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
      \\
    \dnt{\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      ; \dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger
      \\
    \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
      : \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
      \\
    \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
      \\
    \dnt{\lhaslb{\ms{L}}{\cdot}{\ms{L}}}
      = \iota_0
  \end{gather*}
  \caption{Semantics for \isotopessa blocks}
  \Description{Semantics for isotope-SSA blocks}
  \label{fig:ssa-block-semantics}
\end{figure}

\TODO{\textit{what} is the semantics of a block}

We can now give a semantics for blocks in Figure \ref{fig:ssa-block-semantics}.
Our semantics for break statements \(\lbrb{\ell}{a}\) is given by the semantics
for splitting the context into variables which pass into dominated blocks and
variables used to compute the argument, \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\),
with the latter passed into the semantics of the argument (which is pure, and
therefore must be lifted to \(\mc{C}_0\))
\(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result along with the
preserved variables passed into \(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to
obtain the desired target. The semantics of if-statements are given by first
splitting the context via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) and then
calculating the discriminant \(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\),
which must be pure and therefore needs to be lifted. This is then passed through
the distributor \(\delta\), with each branch passed to the semantics of its
corresponding block \(\dnt{\haslb{\Xi}{s}{\ms{L}}}\),
\(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The semantics of \rle{let-blk} and
\rle{let2-blk} are analogous to those of \rle{let} and \rle{let2} in the
expression case, except that now the bound expression \(a\) is allowed to be
impure (and therefore in particular does not need to be upgraded).

@@

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{hole-blk}}
      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
      {\cwk{\Gamma}{\Delta}}
      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
      \qquad
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{\tint}}
      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
      \\
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhasty{H}{\Delta}{1}{\mhole{e}}{\mb{2}}{\tint}}
      {\mhaslb{H}{\Xi}{\mhole{s}}{\ms{L}}}
      {\mhaslb{H}{\Xi}{\mhole{t}}{\ms{L}}}
      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{q}}
      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A \otimes B}{q}}
      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{where}}
      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
      \qquad
    \prftree[r]{\rle{hole-cfg}}
      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
      {\lwk{\ms{M}}{\ms{K}}}
      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
      \\
    \prftree[r]{\rle{nil-br}}
      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks with holes}
  \Description{Typing rules for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-typing}
\end{figure}

>>>>

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      {\hasty{\Xi}{1}{a}{A}{\tint}}
      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} 
    \\
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\hasty{\Xi}{1}{e}{\mb{2}}{\tint}}
      {\haslb{\Delta}{s}{\ms{L}}}
      {\haslb{\Delta}{t}{\ms{L}}}
      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} 
    \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{0}{a}{A}{q}}
      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} 
    \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{0}{a}{A \otimes B}{q}}
      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \qquad
    \prftree[r]{\rle{where}}
      {\haslb{\Gamma}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L}{\ms{K}}}
      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
    \\
    \prftree[r]{\rle{nil-br}}
      {\lhaslb{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks}
  \Description{Typing rules for isotope-SSA blocks}
  \label{fig:ssa-block-typing}
\end{figure}

@@

\begin{figure}
  \begin{gather*}
    \boxed{\dnt{\haslb{\Gamma}{t}{\ms{L}}}
      : \mc{C}_0^\varnothing(\dnt{\Gamma}, \dnt{\ms{L}})} \\
    \dnt{\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}}
      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{a}{A}{\tint}})}
      ; \dnt{\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}} \\
    \dnt{\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}}
      = \upg{(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Xi}{1}{e}{\mb{2}}{\tint}})}
      ; \delta ;
      \\ \qquad \qquad \qquad \qquad
      [
        \dnt{\haslb{\Delta}{s}{\ms{L}}}, 
        \dnt{\haslb{\Delta}{t}{\ms{L}}} 
      ]
      \\
    \dnt{\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}}
      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A}{q}}
      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
      \\
    \dnt{\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}}
      = \upg{\dnt{\csplits{\Gamma}{\Delta}{\Xi}}}
      ; \dnt{\Delta} \otimes \dnt{\hasty{\Gamma}{0}{a}{A \otimes B}{q}}
      ; \alpha
      ; \dnt{\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
      \\
    \dnt{\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      ; \dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}^\dagger
      \\
    \boxed{\dnt{\lhaslb{\ms{L}}{L}{\ms{K}}}
      : \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}} + \dnt{\ms{L}})}
      \\
    \dnt{\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
      = \dnt{\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      ; [[\iota_0, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}], \iota_1]
      \\
    \dnt{\lhaslb{\ms{L}}{\cdot}{\ms{L}}}
      = \iota_0
  \end{gather*}
  \caption{Semantics for \isotopessa blocks}
  \Description{Semantics for isotope-SSA blocks}
  \label{fig:ssa-block-semantics}
\end{figure}

Similarly to terms, \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) are
interpreted as \textit{impure} morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\):
here \(\Gamma\) represents all the variables live on entry to the block, while
\(\ms{L}\) represents all the external blocks which \(t\) might branch to, each
annotated with their live variables and their arguments; our rules are given in
Figure \ref{fig:ssa-block-semantics}. Our semantics for break statements
\(\lbrb{\ell}{a}\) is as follows: we begin by splitting the context into live
variables which pass into dominated blocks and variables used in the argument
via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\). We then pass the latter into the
semantics of the argument (which is pure, and therefore must be lifted to
\(\mc{C}_0\)) \(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result and live
variables re-associated and passed into
\(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target. The
semantics of if-statements are similarly given by first splitting the context
via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) into variables live in the branches,
\(\Delta\), and variables appearing in the discriminant, \(\Xi\); the latter of
which are passed into the lifted pure morphism
\(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\). The "if" statement itself is then
simply an application of the distributor \(\delta: A \otimes (\mb{1} + \mb{1})
\to A + A\), with each branch passed to the semantics of its corresponding block
\(\dnt{\haslb{\Xi}{s}{\ms{L}}}\), \(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The
semantics of \rle{let-blk} and \rle{let2-blk} are analogous to those of
\rle{let} and \rle{let2} in the expression case, except that now the bound
expression \(a\) is allowed to be impure (and therefore in particular does not
need to be upgraded).

@@

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{hole-blk}}
      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
      {\cwk{\Gamma}{\Delta}}
      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
      \qquad
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{\tint}}
      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
      \\
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhasty{H}{\Xi}{1}{\mhole{e}}{\mb{2}}{\tint}}
      {\mhaslb{H}{\Delta}{\mhole{s}}{\ms{L}}}
      {\mhaslb{H}{\Delta}{\mhole{t}}{\ms{L}}}
      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{q}}
      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A \otimes B}{q}}
      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{where}}
      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
      \qquad
    \prftree[r]{\rle{hole-cfg}}
      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
      {\lwk{\ms{M}}{\ms{K}}}
      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
      \\
    \prftree[r]{\rle{nil-br}}
      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks with holes}
  \Description{Typing rules for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-typing}
\end{figure}
